Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-11 paper/main.tex
fixes to expand and fixcycles

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. These
include constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite this significant body of work,
the adoption of causal discovery methods in observational research has so far 
been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the standard
deviation of $ X $ as $ \sigma_X $, the covariance between variables $ X $ and
$ Y $ as $ \mathrm{cov}(X, Y) $, a covariance matrix as $ \Sigma $, and the
entry corresponding to the covariance between $ X $ and $ Y $ as $ \Sigma_{XY}
$. A DAG $ G =
(V, E) $ is an acyclic directed graph whose nodes $ V $ corresponding to random
variables and whose edges $ E $ represent direct causal relationships
\cite{Pearl2009}. For an edge $e=X \to Y$, $\textrm{reverse}(e)=Y \to X$. The
set of parents of $ X $ in $ G $ is denoted as $ \textrm{Pa}_G(X) $, and its
ancestors and descendants as $ \textrm{An}_G(X) $ and $ \textrm{De}_G(X) $,
respectively; we use the convention that $X \notin \textrm{An}_G(X)$ and $X
\notin \textrm{De}_G(X)$. We denote a path, $ \pi(X, Y) = \{ X, V_0, V_1,
\cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G $ where consecutive pairs of
variables in $ \pi $ are connected by an edge. We say $ X $ and $ Y $ are
d-connected in $ G $ given a conditioning set $ \bm{Z} \subseteq V - \{X, Y\}
$, if for a path $ \pi(X, Y) $ (i) for every collider structure ($ V_1
\rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2,
\textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider
structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2
\rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated
given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    $L \gets L \cup \{ e \} $ \;
	    \If{ $ (V, E \cup L) $ is cyclic }{
		    $ L' \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G) $ \;
		    $ B \gets B \cup (L \setminus L') $ \;
		    $ L \gets L'$
	    }
	    % \If{ $(V,E \cup L \cup e)$ is acyclic }{
            % 	$L \gets L \cup  \{ e \} $\;
	    % } \Else {
            % 	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    % }
        }
	\If{$|L|\geq k$}{\KwGoTo 18\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E,B)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ X \rightarrow Y $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(X, Y, \mathbf{Z})=1$
}{
			$ R \gets R \cup \{ X \rightarrow Y \} $ \;

		}
	}
	\Return $ L \setminus R $ \;
}
\caption{Fixing cycles by removing incorrect edges.}
\label{algo:fix_cycles}
\end{algorithm}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$\phi$,$B=\emptyset$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ e_{\max} \gets \mathcal{A}_G(X, Y)$ \;
		}
        }
    }

>>>>

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. These
include constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}, score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite this significant body of work,
the adoption of causal discovery methods in observational research has so far 
been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the standard
deviation of $ X $ as $ \sigma_X $, the covariance between variables $ X $ and
$ Y $ as $ \mathrm{cov}(X, Y) $, a covariance matrix as $ \Sigma $, and the
entry corresponding to the covariance between $ X $ and $ Y $ as $ \Sigma_{XY}
$. A DAG $ G =
(V, E) $ is an acyclic directed graph whose nodes $ V $ corresponding to random
variables and whose edges $ E $ represent direct causal relationships
\citep{Pearl2009}.  The
set of parents of $ X $ in $ G $ is denoted as $ \textrm{Pa}_G(X) $, and its
ancestors and descendants as $ \textrm{An}_G(X) $ and $ \textrm{De}_G(X) $,
respectively; we use the convention that $X \notin \textrm{An}_G(X)$ and $X
\notin \textrm{De}_G(X)$. We denote a path, $ \pi(X, Y) = \{ X, V_0, V_1,
\cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G $ where consecutive pairs of
variables in $ \pi $ are connected by an edge. We say $ X $ and $ Y $ are
d-connected in $ G $ given a conditioning set $ \bm{Z} \subseteq V - \{X, Y\}
$, if for a path $ \pi(X, Y) $ (i) for every collider structure ($ V_1
\rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $ \bm{Z} \cap \{ V_2,
\textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every non-collider
structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1 \leftarrow V_2
\rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are d-separated
given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E \cup B $ and $Y \to X \notin E \cup B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $L \gets L \cup \mathcal{A}_G(X, Y) $ \;
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    $ R \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G) $ \; 
    $ B \gets B \cup R \; ; \; L \gets L \setminus R \; ; \;  E \gets (E \cup L) \setminus B$ \;
    \Return{$(V,E,B)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycles}{FixCycles}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, \mathcal{D}_G $)}{
		$R \gets \emptyset$ \; 
		\ForEach {$ X \rightarrow Y $ on a cycle in $ (V,E) $}{
		\If {there exists a set $\mathbf{Z}\subseteq\mathbf{X}$ where $\mathcal{D}_G(X, Y, \mathbf{Z})=1$
}{
			$ R \gets R \cup \{ X \to Y \} \cup \{ Y \to X \} $ \;

		}
	}
	\Return $ R $ \;
}
\caption{Fixing cycles by removing incorrect edges.}
\label{algo:fix_cycles}
\end{algorithm}

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$\phi$,$B=\emptyset$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E \setminus B$ and $Y \to X \notin E \setminus B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ e_{\max} \gets \mathcal{A}_G(X, Y)$ \;
		}
        }
    }
    $L' \gets \textsc{FixCycles}(V, E \cup L, \mathcal{D}_G)$ \;
